Nowcasting Covid-19 statistics reported with delay: a case-study of Sweden

### 2021-03-03 output/paper/appendix.tex
Updates from Overleaf

@@

\subsection{Priors}
For the $\mv{\alpha}$ and $\mv{\beta}$ parameters we use an (improper) uniform prior. For the deaths, $\mv{d}$, one could imagine several different prior ideally some sort of epidemiological model. However, here we just assume a log-Gaussian Cox processes \citep{Moller1998_log_gaussian}, but instead of Poisson distribution we use a negative binomial to handle possible over dispersion. The latent Gaussian processes has a intrinsic random walk distribution \citep{Rue2005_gaussian_markov} i.e.
\begin{align*}
\log(\mu_i) - \log(\mu_{i-1}) &\sim N(0,\sigma^2),\\
d_i| \lambda_i  &\sim NegBin(\mu_i, \phi).
\end{align*}
This model is created to create a temporal smoothing between the reported deaths.
For the hyperparameter $\sigma^2$ we impose a inverse Gamma distribution, this prior is suitable here because it guarantees that the process is not constant ($\sigma^2=0$) which we know is not the case.
\subsection{Full model}
Putting the likelihood and priors together we get the following hierarchical Bayesian model
\begin{align*}
\sigma^2 &\sim \Gamma(1,0.01) \\
\phi &\sim \Gamma(1,0.01) \\
\alpha_k &\sim U[0,\infty] \\
\beta_k &\sim U[0,\infty] \\
\alpha_k^H &\sim U[0,\infty] \\
\beta_k^H &\sim U[0,\infty] \\
\log(\mu_i) - \log(\mu{i-1}) &\sim N(0,\sigma^2)\\
d_i| \lambda_i  &\sim NegBin(\mu_i,\phi) \\
p_{i,j}|  \mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H &\sim Beta(\alpha^H_j\alpha_{min(j-i,k)},\beta^H_j\beta_{min(j-i,k)}) \\
r_{i,j}|d_i,\mv{r}_{1,1:j},p &\sim \pi Bin(d_i - \sum_{k=1}^{j-1}r_{i,k},p_0)+ (1-\pi) Bin(d_i - \sum_{k=1}^{j-1}r_{i,k}, p_{i,j}),
\end{align*}
where where and $j\leq i$ and $i=1,\ldots,T$.

\section{Inference}
As the main goal to generate inference of the number of death $\mv{d}$ is through the posterior distribution of number of deaths $\mv{d}$ given the observations $\mv{r}$.
In order to generate samples from this distribution we use a Markov Chain Monte Carlo method \citep{Brooks2011_handbook_markov}. In more detail we use a blocked Gibbs sampler, which generates samples in the following sequence:
\begin{itemize}
	\item  We sample $\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H|\mv{d}, \mv{r}$ using the fact that one can integrate out $p$ in the model, and then  $\mv{d}|\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H,\mv{r},\mv{\lambda}$  follows a Beta-Binomial distribution. Here to we use an adaptive MALA \citep{Atchade2006_adaptive_version} to sample from these parameters.
	\item  To sample $\mv{d}|\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H,\mv{r},\mv{\lambda}$, that each death, $d_i$ is conditionally independent, and we just use a Metropolis Hastings random walk to sample each one.
	\item To sample $\mv{\lambda} | \mv{d},\sigma^2$ we again use an adaptive MALA.
	\item Finally We sample $\sigma^2|\mv{d}$,and  $p_0,\pi$ directly since this distribution is explicit, and $\phi$ using a MH-RW.
\end{itemize}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{../plots/SCRPS_over_states}
    \caption{Average SCRPS as the pandemic progresses.}
    \label{fig:SCRPS_states}
\end{figure}

>>>>

\renewcommand{\thefigure}{A\arabic{figure}}
\setcounter{figure}{0}

@@

\subsection{Priors}
For the $\mv{\alpha}$ and $\mv{\beta}$ parameters we use an (improper) uniform prior. For the deaths, $\mv{d}$, one could imagine several different prior ideally some sort of epidemiological model. However, here we just assume a log-Gaussian Cox processes \citep{Moller1998_log_gaussian}, but instead of Poisson distribution we use a negative binomial to handle possible over dispersion. The latent Gaussian processes has a intrinsic random walk distribution \citep{Rue2005_gaussian_markov} i.e.
\begin{align*}
\log(\mu_i) - \log(\mu_{i-1}) &\sim N(0,\sigma^2),\\
d_i| \phi, \mu_i  &\sim NegBin(\mu_i, \phi).
\end{align*}
This model is created to create a temporal smoothing between the reported deaths.
For the hyperparameter $\sigma^2$ we impose a inverse Gamma distribution, this prior is suitable here because it guarantees that the process is not constant ($\sigma^2=0$) which we know is not the case.
\subsection{Full model}
Putting the likelihood and priors together we get the following hierarchical Bayesian model
\begin{align*}
\sigma^2 &\sim \Gamma(1,0.01) \\
\phi &\sim \Gamma(1,0.01) \\
\alpha_k &\sim U[0,\infty] \\
\beta_k &\sim U[0,\infty] \\
\alpha_k^H &\sim U[0,\infty] \\
\beta_k^H &\sim U[0,\infty] \\
\log(\mu_i) - \log(\mu_{i-1}) &\sim N(0,\sigma^2)\\
d_i| \phi,\mu_i  &\sim NegBin(\mu_i,\phi) \\
p_{i,j}|  \mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H &\sim Beta(\alpha^H_j\alpha_{min(j-i,k)},\beta^H_j\beta_{min(j-i,k)}) \\
r_{i,j}|d_i,\mv{r}_{1,1:j},p &\sim \pi Bin(d_i - \sum_{k=1}^{j-1}r_{i,k},p_0)+ (1-\pi) Bin(d_i - \sum_{k=1}^{j-1}r_{i,k}, p_{i,j}),
\end{align*}
where where and $j\leq i$ and $i=1,\ldots,T$.

\section{Inference}
As the main goal to generate inference of the number of death $\mv{d}$ is through the posterior distribution of number of deaths $\mv{d}$ given the observations $\mv{r}$.
In order to generate samples from this distribution we use a Markov Chain Monte Carlo method \citep{Brooks2011_handbook_markov}. In more detail we use a blocked Gibbs sampler, which generates samples in the following sequence:
\begin{itemize}
	\item  We sample $\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H|\mv{d}, \mv{r}$ using the fact that one can integrate out $p$ in the model, and then  $\mv{d}|\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H,\mv{r},\mv{\mu},\phi$  follows a Beta-Binomial distribution. Here to we use an adaptive MALA \citep{Atchade2006_adaptive_version} to sample from these parameters.
	\item  To sample $\mv{d}|\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H,\mv{r},\mv{\mu},\phi$, that each death, $d_i$ is conditionally independent, and we just use a Metropolis Hastings random walk to sample each one.
	\item To sample $\mv{\mu} | \mv{d},\sigma^2,\phi$ we again use an adaptive MALA.
	\item Finally We sample $\sigma^2|\mv{d}$,and  $p_0,\pi$ directly since this distribution is explicit, and $\phi$ using a MH-RW.
\end{itemize}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{../plots/SCRPS_over_states}
    \caption{Average SCRPS as the pandemic progresses.w}
    \label{fig:SCRPS_states}
\end{figure}

### 2021-03-03 output/paper/paper.tex
Updates from Overleaf

\documentclass[a4paper,11pth]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
%\usepackage{babel}
%\usepackage[protrusion=true,expansion=true,final,babel]{microtype}
%\usepackage[babel]{csquotes}
%\usepackage{babel}
\usepackage{natbib}
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{booktabs}
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf,.png,.jpg,.eps}
\usepackage{array}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{hyperref}
\newcolumntype{L}{>{\raggedright\arraybackslash}X}
\newcommand{\mv}[1]{{\boldsymbol{#1}}}

@@

\subsection{The current situation of COVID-19 in Sweden}
The Swedish Public Health Agency updates the COVID-19 statistics daily\footnote{The data is published on \url{https://www.folkhalsomyndigheten.se/smittskydd-beredskap/utbrott/aktuella-utbrott/covid-19/bekraftade-fall-i-sverige/}.}. During frequently re-occurring press conference, they present updates on the number of deaths, admissions to hospitals and intensive care, as well as case counts.

@@

\section{The removal method}
To correct for the reporting delays we propose to use the removal method, developed in animal management \citep{Pollock1991_review_papers}, to present an estimate of the actual frequencies at a given day and their uncertainty. The method has a long history dating back at least to the 1930s \citep{Leslie1939_attempt_determine}. However, the first refined mathematical treatment of the method is credited to \citet{Moran1951_mathematical_theory}, more modern derivatives exits today \citep{Matechou2016_open_models}. It is a commonly applied method today when analyzing age cohorts in fishery and wildlife management.

@@

We note that a similar method has been proposed for delay adjusted reporting, but without using the removal method \cite{lawless1994adjustments}. In \cite{lawless1994adjustments} the author developed the Occurred But Not Reported method (OBNR), which can give a frequentist point estimate of the number of deaths, and under a Poisson distributional assumption, generate prediction intervals. A Bayesian version of this method for nowcasting was developed in \cite{hohle2014bayesian}.

@@

% Author contributions
\bibliographystyle{chicago}
\bibliography{bibliography}
\input{contributions.tex}%

>>>>

\documentclass[a4paper,11pth]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[square,numbers]{natbib}
%\usepackage[style=nature]{biblatex}
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{booktabs}
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf,.png,.jpg,.eps}
\usepackage{array}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{hyperref}
\newcolumntype{L}{>{\raggedright\arraybackslash}X}
\newcommand{\mv}[1]{{\boldsymbol{#1}}}

@@

\subsection{The current situation of COVID-19 in Sweden}
The Swedish Public Health Agency updates the COVID-19 statistics daily (the data is published on https://www.folkhalsomyndigheten.se/smittskydd-beredskap
/utbrott/aktuella-utbrott/covid-19/bekraftade-fall-i-sverige/).. During frequently re-occurring press conference, they present updates on the number of deaths, admissions to hospitals and intensive care, as well as case counts.

@@

\section{The removal method}
To correct for the reporting delays we propose to use the removal method, developed in animal management \citep{Pollock1991_review_papers}, to present an estimate of the actual frequencies at a given day and their uncertainty. The method has a long history dating back at least to the 1930s \citep{Leslie1939_attempt_determine}. However, the first refined mathematical treatment of the method is credited to \cite{Moran1951_mathematical_theory}, more modern derivatives exits today \citep{Matechou2016_open_models}. It is a commonly applied method today when analyzing age cohorts in fishery and wildlife management.

@@

We note that a similar method has been proposed for delay adjusted reporting, but without using the removal method \citep{lawless1994adjustments}. In \cite{lawless1994adjustments} the author developed the Occurred But Not Reported method (OBNR), which can give a frequentist point estimate of the number of deaths, and under a Poisson distributional assumption, generate prediction intervals. A Bayesian version of this method for nowcasting was developed in \cite{hohle2014bayesian}.

@@

% Author contributions
\bibliographystyle{unsrt}
\bibliography{bibliography}
\input{contributions.tex}%

\newpage
\section{Figure legends}
Figure 1. Swedish Covid-19 deaths as of 2020-06-04 and model predictions.

Figure 2. Model accuracy over time for four dates, compared to a benchmark.

Figure 3. Model accuracy over time for four dates, compared to benchmark.
